An open and continuously updated fern tree of life (FTOL)

### 2022-02-09 ms/manuscript.Rmd
Update MS
use initial time tree

```{r load-targets}
# Targets from FTOL workflow
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_tree,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl),
	store = here::here("ftol_cache")
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files
	),
	store = here::here("_targets")
)

# FIXME: read in real ultrametric tree when ready
sanger_tree <- ape::read.tree(here::here("working/sanger_alignment.phy.tre"))
```

@@

We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013]. 
### Initial Sanger phylogeny

Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the same model as selected for the backbone phylogeny.
We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.

@@

We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure as used for the backbone phylogeny.
Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = plastome_tree$tip.label) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		plot_group = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ plot_group)
	)
```

@@

genbank_plot <-
	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
	geom_line(aes(color = type)) +
	scale_color_manual(values = compartment_cols, guide = "none") +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("nuclear", "mitochondrial")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
    # Do not repel from top or bottom edges.
    ylim = c(-Inf, Inf)
	) +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("total", "plastid")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
    # Do not repel from top or bottom edges.
    ylim = c(-Inf, Inf),
		min.segment.length = 0
	) +
	geom_point(
		data = fern_tree_sampling,
		aes(shape = source)
	) +
	scale_shape(
    limits = fern_tree_sampling$source, 
    labels = fern_tree_sampling$label
	) +
	labs(shape = "Study", y = "No. species", x = "Year") +
	theme(legend.position = "bottom",
				plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")) +
	guides(shape = guide_legend(title.position = "top")) +
	coord_cartesian(clip = "off")
```

@@

# Make exemplar set of species in backbone representing plot groups
plot_group_exemp_taxa <-
	backbone_taxonomy %>%
	# Slice to one exemplar species per family
	group_by(plot_group) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order)

# Get names of exemplar tips in ladderized phylo order
# Trim tree to exemplar species, rename by family
phy <- ape::keep.tip(plastome_tree, plot_group_exemp_taxa$species)
phy$tip.label <- plot_group_exemp_taxa$plot_group

# Make tibble of plot groups in phylogenetic order
plot_group_phy_order <-
	phy %>%
	# Reroot, ladderize, drop outgroup
	root.phylo("og") %>%
	ladderize() %>%
	drop.tip("og") %>%
	jntools::get_tips_in_ape_plot_order() %>%
	tibble(
		name = .,
		phy_order = seq_along(.)
	)

rm(phy)

@@

clade_cols <-
	Polychrome::green.armytage.colors() %>%
	extract(green_army_use_cols) %>%
	extract(seq_len(nrow(plot_group_phy_order))) %>%
	set_names(plot_group_phy_order$name)

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	left_join(plot_group_phy_order, by = "name") %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_text(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

# Tree plot ----
phy <- sanger_tree %>%
	root("Anthoceros_angustus") %>%
	ladderize()

@@

a <- ggtree(
	phy, 
	layout = "circular", right = FALSE, size = 0.1,
	# FIXME: change to phylogram when ultrametric tree is done
	branch.length = "none"
) +
	geom_cladelab(
		data = clade_tibble,
		mapping = aes(
			node = mrca,
			color = plot_group,
			label = dummy
		),
		offset = 0.1,
		barsize = 2) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")

>>>>

```{r load-targets}
# Targets from FTOL workflow
tar_load(
	c(
		sanger_alignment,
		plastome_alignment,
		plastome_metadata_renamed,
		ppgi_taxonomy,
		plastid_calibration_dates,
		pteridocat,
		sanger_alignment_tbl,
		sanger_tree_fast,
		plastome_tree,
		plastid_tree_dated),
	# FIXME: change to ftol_cache when ready
	store = here::here("working/_targets")
)

# Targets from MS workflow
tar_load(
	c(
		gb_species_by_year,
		ref_files
	),
	store = here::here("_targets")
)
```

@@

We generated a backbone phylogeny using maximum likelihood (ML) analysis of the concatenated plastome dataset in IQ-TREE [@Nguyen2015].
ModelFinder [@Kalyaanamoorthy2017] was implemented in IQ-TREE to select the best-fitting model of sequence evolution.
To reduce computational burden, we only tested models based on the General Time Reversible (GTR) model [@Tavare1986] and did not partition the dataset.
The best-fitting model was selected automatically by IQ-TREE according to Bayesian Information Criterion (BIC) score.
Node support was assessed with `r number(1000)` ultrafast rapid bootstrap replicates [@Minh2013].

### Initial Sanger phylogeny

Next, we used the backbone phylogeny as a constraint tree to conduct initial phylogenetic analysis of the concatenated Sanger dataset in IQ-TREE with the "-fast" option and the GTR+I+G model.
We prioritized computation speed for this step, as we needed to repeat it multiple times as described in the next section.

@@

We generated the final Sanger phylogeny by using the backbone phylogeny as a constraint tree in ML analysis of the final concatenated Sanger dataset with the same model selection procedure as used for the backbone phylogeny.
Node support was assessed with `r number(100)` standard (non-parametric) bootstrap replicates.

@@

plastome_tree_rooted <- phytools::reroot(
	plastome_tree,
	getMRCA(
		plastome_tree,
		c("Physcomitrium_patens", "Marchantia_polymorpha", "Anthoceros_angustus")
	)) %>%
	ladderize()

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(
		species = get_tips_in_ape_plot_order(plastome_tree_rooted)
	) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		plot_group = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ plot_group)
	)
```

@@

# Manually specify axis buffers since we will annotate the panel border manually
year_buffer <- 
	max(gb_species_by_year$year) %>%
	subtract(min(gb_species_by_year$year)) %>%
	multiply_by(0.05)
sp_buffer <- max(gb_species_by_year$n_species) %>%
	subtract(min(gb_species_by_year$n_species)) %>%
	multiply_by(0.05)

genbank_plot <-
	ggplot(gb_species_by_year, aes(x = year, y = n_species)) +
	geom_line(aes(color = type)) +
	scale_color_manual(values = compartment_cols, guide = "none") +
	# Annotate panel border manually, since the automatic border gets
	# printed on top of label annotations with theme_bw()
	annotate(
		"rect", 
		xmin = min(gb_species_by_year$year) - year_buffer,
		xmax = max(gb_species_by_year$year) + year_buffer,
		ymin = min(gb_species_by_year$n_species) - sp_buffer,
		ymax = max(gb_species_by_year$n_species) + sp_buffer,
		fill = "transparent",
		color = "grey20") +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("nuclear", "mitochondrial")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		# Do not repel from top or bottom edges.
		ylim = c(-Inf, Inf)
	) +
	geom_label_repel(
		data = compartment_labs %>% filter(type %in% c("total", "plastid")),
		aes(color = type, label = label),
		segment.linetype = 2, 
		xlim = c(2020, Inf),
		ylim = c(-Inf, Inf),
		min.segment.length = 0
	) +
	geom_point(
		data = fern_tree_sampling,
		aes(shape = source)
	) +
	scale_shape(
		limits = fern_tree_sampling$source, 
		labels = fern_tree_sampling$label
	) +
	scale_x_continuous(expand = expansion(0, 0)) + 
	scale_y_continuous(expand = expansion(0, 0)) + 
	labs(shape = "Study", y = "No. species", x = "Year") +
	coord_cartesian(clip = "off") +
	guides(
		shape = guide_legend(
			title.position = "top",
			nrow = 2)
	) +
	theme_bw() +
	theme(
		panel.border = element_blank(),
		legend.position = "bottom",
		plot.margin = margin(r = 1, t = 0.5, l = 0.5, unit = "in")
	)
```

@@

# Make exemplar set of species in backbone representing plot groups
plot_group_exemp_taxa <-
	backbone_taxonomy %>%
	# Slice to one exemplar species per major clade
	group_by(plot_group) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order) %>%
	# Drop outgroup
	filter(family != "og")

@@

clade_cols <-
	Polychrome::green.armytage.colors() %>%
	extract(green_army_use_cols) %>%
	extract(seq_len(nrow(plot_group_exemp_taxa))) %>%
	set_names(plot_group_exemp_taxa$plot_group)

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	left_join(plot_group_exemp_taxa, by = c(name = "plot_group")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_text(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

# Tree plot ----
phy <- plastid_tree_dated

@@

a <- ggtree(
	phy, 
	layout = "circular", right = FALSE, size = 0.1,
) +
	geom_cladelab(
		data = clade_tibble,
		mapping = aes(
			node = mrca,
			color = plot_group,
			label = dummy
		),
		offset = 0.1,
		barsize = 2) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")
